Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome

### 2023-06-01 paper/paper.qmd
Introduce NNB

When considering the suitability of a model for deployment in clinical settings, 
the number needed to screen (NNS) is a highly relevant metric representing how
many patients must be predicted as severe by the model to identify one true positive.
We determined the decision threshold at the 95th percentile
of risk for each prediction model and computed the confusion matrix and several
performance metrics at this threshold including the NNS (@tbl-risk).
Among the models predicting severe outcomes, the models trained on the full
datasets performed best with an NNS of `r allcause_nns` for the all-cause
definition, `r attrib_nns` for the attributable definition, and 
`r pragmatic_nns` for the pragmatic definition.
For context, prior studies predicted CDI-attributable severity using whole
Electronic Health Record data and from a smaller set of 9 clinician-curated
variables, achieving precision values of 0.417 (NNS = `r 1/0.417`) for the EHR
model and 0.167 (NNS = `r 1/0.167`) for the curated model [@li_using_2019;
@rao_clostridium_2015].
While the attributable definition had a worse NNS for our OTU-based models, it
did not perform worse than prior curated models and it is the most clinically
relevant as physician chart review ensures that positively-labelled severe
outcomes are due to the CDI rather than other causes.

Even if a model performs well, it may not be useful in a clinical setting unless
it can guide clinicians to choose between treatment options.
At this time, we are not aware of any direct evidence that a particular
treatment reduces the risk of severe CDI outcomes.
However, with some assumptions we offer a proof-of-concept analysis of the
potential clinical value of OTU-based severity prediction models.
Paired with the NNS of these models,

Current clinical guidelines specify vancomycin and fidaxomicin as the standard
antibiotics to treat CDI, with a preference for fidaxomicin due to its higher
sustained resolution of CDI and reduced risk of recurrence
[@stuart_clinical_2021].
However, fidaxomicin is considerably more expensive than vancomycin, resulting
in most CDI patients in the US being treated with vancomycin (TODO citation).
If fidaxomicin were shown to reduce the risk of severe CDI outcomes, it could be
preferentially prescribed to patients predicted to be at risk.
While we are not aware of evidence demonstrating the superiority of fidaxomicin
for reducing severe CDI outcomes,

@@

It's not enough for models to perform well to justify deploying them in a
clinical setting; benefit over current practices must be shown.
Amplicon sequencing is not typically performed for CDI patients, but if there is
clinical value to be gained by implementing OTU-based models, routinely
sequencing and profiling the microbial communities of CDI patients could be
justified.

>>>>

Even if a model performs well, it may not be useful in a clinical setting unless
it can guide clinicians to choose between treatment options.
At this time, we are not aware of any direct evidence that a particular
treatment reduces the risk of severe CDI outcomes.
However, with some assumptions we offer a proof-of-concept analysis of the
potential clinical value of OTU-based severity prediction models when paired
with treatments that may reduce severity.
When considering the suitability of a model for deployment in clinical settings,
the number needed to screen (NNS) is a highly relevant metric representing how
many patients must be predicted as severe by the model to identify one true
positive.
The number needed to treat (NNT) is the number of true positive patients that
must be treated by an intervention in order for one patient to benefit from the
treatment. 
Multiplying NNS by NNT yields the number needed to benefit (NNB): the number
of patients predicted to have a severe outcome who then benefit from the
treatment [@liu_number_2019].
The NNB pairs model performance with treatment effectiveness to estimate the
benefit of using predictive models in clinical practice.

We determined the decision threshold at the 95th percentile of risk for each
prediction model and computed the confusion matrix and several performance
metrics including the NNS (@tbl-risk).
Among the models predicting severe outcomes, those trained on the full
datasets performed best with an NNS of `r allcause_nns` for the all-cause
definition, `r attrib_nns` for the attributable definition, and 
`r pragmatic_nns` for the pragmatic definition.
For context, prior studies predicted CDI-attributable severity using whole
Electronic Health Record data and from a smaller set of clinician-curated
variables, achieving precision values of 0.417 (NNS = `r 1/0.417`) for the EHR
model and 0.167 (NNS = `r 1/0.167`) for the curated model [@li_using_2019;
@rao_clostridium_2015].
While the attributable definition had a worse NNS for our OTU-based models, it
did not perform worse than prior curated models and it is the most clinically
relevant as physician chart review increases confidence that positively-labelled
severe outcomes are due to the CDI rather than other causes.

Current clinical guidelines specify vancomycin and fidaxomicin as the standard
antibiotics to treat CDI, with a preference for fidaxomicin due to its higher
sustained resolution of CDI and reduced risk of recurrence
[@stuart_clinical_2021].
However, fidaxomicin is considerably more expensive than vancomycin, resulting
in most CDI patients in the US being treated with vancomycin (TODO citation).
If fidaxomicin were shown to reduce the risk of severe CDI outcomes, it could be
preferentially prescribed to patients predicted to be at risk.
While we are not aware of evidence demonstrating the superiority of fidaxomicin
for reducing severe CDI outcomes, the reduced risk of recurrence

@@

It's not enough for models to perform well to justify deploying them in a
clinical setting; benefit over current practices must be shown.
Estimating the NNB contextualizes model performance within clinical reality.
Amplicon sequencing is not typically performed for CDI patients, but if there is
clinical value to be gained by implementing OTU-based models, routinely
sequencing and profiling the microbial communities of CDI patients could be
justified.
